---
bibliography: paper.bib
#cls: style.cls
---
# Temperature-dependence in sewer blockage frequency 
Josh Nightingale, Christian Gunning and Mark Holstad

```{r setup, include=FALSE, cache=FALSE}
## table digits, see mk.helpers
.tab.dig = 3
## long
.do.grid <- FALSE
# recompile will have same random numbers
set.seed(76543); 
## includes opts_chunk
source('init.R')
source('mk.helpers.R')

## Rest of chunk deprecated?
# Code to set up knitr settings and libraries used every time
## 

## set up figure numbers and captions
## knitr options, hooks
#source('prep.knitr.markup.R')

# set global chunk options
## josh: this line never works on my machine and causes compiling to fail...
#opts_chunk$set(fig.align='center', fig.show='hold', concordance=TRUE, tidy=T)
#options(replace.assign=TRUE, width=70)
#setwd('/home/josh/unm-r-programming/sewer')
```

```{r read-data, size='footnotesize', include=FALSE, cache=F}
## this also sources run.model_sew_temp.R and run.load_fog.R
source('run.read_data.R')
## read in FOG data, and some processing
source('run.load_fog.R')

## find best timeperiod to integrate air temp over
## model best N days for rolling mean
source('run.model_sew_temp.R')
## join, cleanup (uses rolling mean above)
source('run.combine.data.R')
```

```{r run-mods, size='footnotesize', include=FALSE}
## 2D gaussian
if (.do.grid) {
    source('run.grid.R')
}
## prepare linear models / diagnosis plots
## was source('sewtempandblock.R')
source('run.glms.R')
```

## Introduction
Sanitary sewer blockages (SSB) cause widespread negative impacts, including
aesthetic degradation from odors, and property damage and environmental
degradation from sanitary sewage overflow (SSO).  In the U.S., SSOs are
tracked by the Environmental Protection Agency [@epa2004local]. Here, approximately
half of SSOs were caused by blockages, with up to 75\% of SSOs caused by
blockages in the arid Southwest [@epa2004report].  Consequently, 
the prompt remediation of SSB is a high priority for municipalities, 
and plays a major role in sanitary sewer maintenance 
costs [@epa2004report, fenner2000approaches].

Here we use an extensive dataset of SSB events in Albuquerque, New Mexico to
demonstrate a significant association between blockage frequency and air
temperature.  We find that, at suitable time lags, air temperature is a 
very good predictor of sewer temperatures. We demonstrate that low
environmental temperatures are 
followed by modest increases in SSB frequency. We discuss the physical and
operational significance of these findings, and their applications to other 
localities.

### Drivers of Blockage

An extensive body of literature exists on SSB mechanism, 
physical structure, and 
detection [@fenner2000approaches, @Keener2008, @williams2012fat, @dominic2013understanding @he2013mechanisms].
In addition, several large-scale studies have estimated drivers of SSB 
rates at the system level [@Marlow2011, @mattsson2014fat].

Previous work has attributed SSBs primarily to roots, debris, and fats, oils,
and grease (FOG) [@epa2004report].  In the U.S., 60-75\% of blockages have 
FOG deposits as a contributory factor [@Keener2008], while
vegetation intrusion is the chief cause of blockages in Australia
[@Marlow2011].  

[](Surprisingly rare, however, are large-scale studies of the correlates and putative causes of
elevated SSB rates. )


In general, higher blockage frequencies are reported in systems with lower
average water use, likely due to increased solid concentration
[@marleni2012impact].  Weather and climate can also influence SSB rates via
affects on both vegetation and water flow.  @Marlow2011, for example, showed a
correlation between sewer blockage frequency and the Southern Oscillation Index
(SOI) in eastern Australia. The SOI reflects rainfall patterns in the region,
with droughts raising blockage risk by decreasing sewer flow volume and
increasing sedimentation. Low rainfall also promotes tree root development,
which damage pipes by intruding through joins and other weak points
[@Desilva2011]. Note, however, that tree root intrusion appears much less
common in systems installed after circa 1970 [@randrup2001tree].

As recognized contributors to SSB, FOG deposits have received considerable
attention. FOG deposits form in a saponification reaction between calcium soaps
and free fatty acids [@He2011], chiefly from restaurants and industrial sources
[ @Keener2008].  Free fatty acids are insoluble in water, and are transported in
greasy effluent.  Many municipalities have implemented policies to minimize FOG
inputs into sanitary sewers [@chu1999pollution, @hassey2001grease, @heckler2003best,
@parnell2005innovative, @bennett2006atlanta, @tupper2008fog, @ducoste2008assessment].  
Residential
outreach is often increased during the holiday season in an effort to minimize
FOG inputs due to food preparation [@tupper2008fog].

### Sewer Struction and Function
Two main forms of sewer systems have been commonly employed 
throughout history: combined sewer systems (CSS) that 
carry both storm water and
sanitary waste in a single sewer network, and separate sewer 
systems (SSS) that use separate networks for each.  CSSs 
were predominant before the late 19^th^ century, with examples of
SSS growing more common in the late 19^th^ and early 20^th^ 
century [@burian2000urban]. In the U.S., rapid population growth,
increased urbanization, and and increased attention to water 
quality throughout the 20^th^ century led to increased use of 
SSS, especially in new developments. SSSs, which we focus on here,
are notable for higher solids loads and more predictable
flow rates. In CSOs, high-flow weather events can help flush 
solids, but can also cause weather-related overflow. In addition,
the types of solids commonly found in SSS versus CSS can differ,
with CSS carrying greater portions of inorganic sand, clay, and 
litter [@arthur2008understanding].

Cultural practices such as diet and FOG management can also influence
SSB frequency through FOG load.
[](FOG discharge rates tend to increase with rising income. ??)
Food service establishments such as restaurants are major 
point-source generators of FOGs in wastewater [@chen2000separation].
Consequently, FOGs discharge from food service establishments
is closely regulated in many locales, including the U.S.,
though FOG interceptor compliance and efficacy can
can vary [@ducoste2008assessment, @williams2012fat].

### Influence of Temperature
Temperature is one potential driver of SSB that has received relatively little 
attention to date. The viscosity of both water and FOGs increases with decreasing
temperature. For a given pipe network, increased viscosity results in increased
frictional head loss [@romeo2002improved]. In addition, FOG effluent
can solidify at lower temperatures, causing overt blockages.

In this study we examine `r length(unique(year(index(block.all.xts))))-1` years 
of SSB records from the City of Albuquerque
municipal sewer system.  We find that air temperature is a useful proxy of
sewage temperature, and that both air and sewage temperature predict SSB
frequency. Specifically, temperature predicts SSB events for which FOGs are a
contributory factor, suggesting that cold weather increases the impact of FOG
deposits. SSBs with other causes do not respond to temperature. 

These relationships shed light on mechanisms
of sewer blockage, and can potentially help municipalities anticipate time
periods of elevated sewer blockages using readily available climatic data.

## Methods

### Data
Albuquerque Bernalillo County Water Utility Authority (ABCWUA) 
covers a 785 km^2^ (303 miles^2^) sewer service area that is roughly bisected by the Rio Grande.
[](The river valley floor itself is nearly flat, with an average river slope of 0.06%). 
The ABCWUA service area includes a population of approximately 606,000 residents located 
primarily within the City of Albuquerque. 
The early system developed in the valley floor and is quite flat, with typical
slopes of 0.3 to 0.5%. Later growth on the east and west mesas has steeper
typical slopes ranging from 0.4 to 3%, with a total elevation change of approximately
460 meters (1,500 feet).  The service area is limited by a 10% slope
demarcation at the foot of the Sandia Mountains.

At present, the system contains more than 182,000 sanitary sewer service
connections. These are serviced by approximately 3,500 km (2,200 miles) of
gravity sewer mains and interceptors, along with 330 km (205 miles) of force and
vacuum lines and 45 lift and vacuum stations. 
Approximately 20% of gravity pipes are greater than 50 years old, while 42% are 
less than 25 years old. The smallest typical gravity line is 0.203 meter (8 inch) pipe, which 
constitutes 82.6% of the gravity mains; the largest pipe in the system is 2.134 meters (84 inches).
Polyvinyl chloride pipe (PVC) constitutes 40% of the system, vitrified
clay pipe (VCP) is 32%, and concrete is 17%, with various other materials
constituting the remainder. 

ABCWUA responds to SSB
events after discovery by maintenance workers or reports of blockages from the
public. This study used an anonymised dataset of SSB dates, along with
engineers' estimates of blockage cause. In total, `r sewer.stats$nobs` SSB
reports from the period `r sewer.stats$min` to `r sewer.stats$max` (inclusive)
were used in this study.  For simplicity, we categorize all reported causes as
either grease related or non-grease related.  The frequency of sanitary sewer
blockages is the primary focus of this work.

As part of an unrelated ABCWUA system odor control survey, sewage temperature
was obtained via grab samples by a chemical treatment vendor at 15 manholes on
three treated north-south interceptors.  A map of these three interceptors is
shown in Figure S1. Temperature measurements were obtained, typically
bi-weekly, using a Cooper Atkins DPP400W Waterproof Pen Style thermometer with
an accuracy range of +/- 1 C° (from -10° to 100°C) in a liquid sample collected
from the manhole.  In total, `r nrow(sewtemp)` sanitary sewer temperature (ST)
measurements from the period `r sewtemp.stats$min` to `r sewtemp.stats$max`
(inclusive) were used in this study.

Mean daily air temperature (mAT/D) was obtained from the Albuquerque
International Airport's (KABQ) automated METAR data collection system
[@abq-metar] spanning the entire period of study noted above.
Direct FOG measurements were available for 
`r length(unique(baw.join$year.week))` weeks. As with temperature,
FOG measurements were averaged to calculate a mean weekly FOG level to use as a predictor
of blockage frequency.

Since most days had no SSB events, the total number of SSB events per week
(SSB/W) was computed and used in subsequent analysis.  For comparison with SSB
data, mAT/D measurements were averaged by week to yield mean weekly air
temperature (mAT/W).  In addition, ST measurements were averaged by week (all
interceptors were combined), yielding mean weekly SGST (mST/W).  In all
analyses that included mST/W, weeks without ST measurements were excluded.  A
time series of SSB events per week, for all causes and grease-related causes,
is shown in Figure \@ref(fig:blockts).

```{r blockts, echo=FALSE, size='footnotesize', include=TRUE, fig.height=4.5, fig.cap="Blockage events per week, for all causes and grease-related causes.  Seasonality of both time series is evident. An exceptional spike of blockages in April 2011 follows a record-breaking cold spell in February 2011.  Likewise, the dearth in blockages in early 2014 and 2015 correspond with historic warm winters in those years."}
xyplot(block.plot.xts, 
    type=c('h','g'), col='black', 
    xlab='', ylab='SSB events per week (SSB/W)',
    par.settings=.lat.theme
)
```

### Linear models
First, we seek to quantify the dependence of sewer temperature on air
temperature using a set of linear models. Exploratory data analysis shows that
manhole identity does not reliably covary with sewage temperature, while both
interceptor identity and air temperature are significant predictors of sewage
temperature.

To account for the lag between air temperature and sewer temperature, we
compute the moving average of air temperature over a varying number of
preceding days N.  For each N, we fit a linear model (using a MANOVA model
structure) that includes interceptor identity and N-day mean air temperature as
predictors.  We then select N to maximize model R^2^.

Next, we seek to quantify the dependence of blockage frequency on temperature.
We model blockage frequency using the Negative Binomial Generalized Linear
Model (NB-GLM) [@venables2013modern].  We start by considering blockages of all
causes, and model their dependence on either sewage temperature or air
temperature.  Due to the sparsity of sewage temperature data, we then focus on
air temperature, and model blockages by both cause and the N-day mean air
temperature. 

Model validation was conducted by subdividing the period of record. We tested
the model of grease blockage frequency's dependence on air temperature, as this
had high predictive value while using a minimal amount of easily-obtained data.
We also test the hypothesis that the Thanksgiving period results in increased
FOG by comparing the model residuals (number of excess blockages not explained
by temperature alone) between weeks containing and immediately after
Thanksgiving with all other weeks during the study period.

Finally, we conduct a detailed analysis of the dependence of grease-caused
blockages on N-day mean air temperature. This analysis involved comparing the
relative utility of temperature as a predictor of blockages with the use of
direct measures of FOG levels, as well as testing the relationship between FOG
measurements, seasonality and air temperature.

Note that standard R^2^ statistics are not available for GLMs.  However, the
proportional reduction in deviance (D) provides an analogous measure of the
model's explanatory power [@zheng2000summarizing]. Thus D falls between 0 and
1, and can be approximately interpreted as the proportion of variance explained
by the model.

All analysis was conducted with the R statistical programming environment [@cran].

## Results

### Mean weekly air temperature (mAT/W) and sewage temperature (mST/W)

Overall, we found that air temperature, averaged over the preceding 
`r best.ndays` days, was a very good predictor of sewage temperature 
(Figure \@ref(fig:sewweathplot)).  Indeed, the final model of sewage temperature versus
air temperature and interceptor identity (Table \@ref(tab:bestlin)) explained
the majority of variation in mST/W (R^2^ = `r round(.best.rsq, digits=3)`).
However, as air temperature falls below freezing, little further decrease in
sewage temperatures was observed (Figure \@ref(fig:sewweathplot)).

We further analysed the effect of local geography on linear model results.  In
all high-ranked linear models, interceptor identity exhibits a small but
statistically significant effect on sewage temperature, while manhole identity
was not a significant predictor.  Nonetheless, the effect size of either
manhole and interceptor identity is small, and we do not consider local
geographic effects further.

### Sewage temperature (mST/W), air temperature (mAT/W), and sewer blockage frequency (SSB/W)

We begin with an examination of all blockages, regardless of reported cause.
We modeled the response of sewer blockage frequency (SSB/W) to either air
temperature (mAT/W) or sewage temperature (mST/W) using a NB-GLM.  Overall, we
found that both sewage temperature and air temperature were weak but
statistically significant predictors of  sewer blockage frequency (Figure
\@ref(fig:multiblockplot), Table \@ref(tab:tempandblock)), For the final
models, we find that D = `r  block.bothtemp.list$devs$Air` (air temperature)
and  D = `r  block.bothtemp.list$devs$Sewer` (sewer temperature). Thus, air
temperature predicts approximately 15% of overall variation in weekly sewer
blockage frequency.   

In the above models, sewer temperature appears to be a better predictor of
blockage frequency than air temperature.  However, the two models are not
directly comparable due to differences in sampling period. Sewage temperature
measurements are expensive and labor-intensive, which in turn limits sample
coverage.  Air temperature records, on the other hand, are freely available
from automated weather stations worldwide, including all major airports.
Consequently, air temperature records cover the entire period of sewer blockage
records.

### Causes of sewer blockage frequency (SSB/W)

Next, we explore the dependence of blockage frequency on reported blockage
cause, grouped into grease versus not grease.  Again, we conduct separate
analyses for air and sewage temperature using a set of NB-GLMs.

This dataset includes `r block.counts$grease` total blockages where grease was
the estimated blockage cause, representing 
`r sprintf("%2.1f", grease.ratio * 100)` percent 
of total incidents during the study period. We constructed a NB-GLM
using both air temperature (mAT/W) and blockage cause (grease vs not grease) as
predictor variables.  Here, both air temperature and blockage cause are highly
significant predictors of blockage frequency (Figure \@ref(fig:block-airtemp)
and Table \@ref(tab:greasetab)). Again, there is a strong interaction between
temperature and blockage cause, with grease-caused blocks responding more
strongly to temperature.   The full model explains a sizable proportion of
variation in blockage frequency (D = `r block.airtemp.list$dev`), particularly
for grease-caused blockages (Figure \@ref(fig:block.airtemp)A).

```{r block-sewtemp, echo=FALSE, fig.pos='h', fig.height=3.5, fig.cap=sprintf('Mean weekly sewer temperature predicts blockages caused by grease (D=%s), and is a poor predictor of blockages due to other causes (D=%s). Models include %d weeks total. See Table ?? for model details.', block.sewtemp.list$dev.grease, block.sewtemp.list$dev.nogrease, block.sewtemp.list$nweeks), message=FALSE}
plot(block.sewtemp.list$plot)
```

### The role of FOG

Blockages were most frequent in weeks which had both high levels of FOG and low
mean air temperatures (Figure \@ref(fig:fogcont)). However, this model
explained a similar amount of deviance (D = `r block.foglevel.list$dev.full`)
to a model containing only temperature and blockage cause as predictors of
blockages during the same weeks (D = `r block.foglevel.list$dev.nofog`). The
addition of FOG to the model resulted in a change of only 
`r block.foglevel.list$aic.nofog - block.foglevel.list$aic.full` AIC units,
whereas an extra parameter should result in a decrease of at least 2 units to
be considered an improvement [@arnold2010uninformative]. Furthermore, FOG level
was not a significant term in this model, nor in models predicting only
grease-caused or other blockages (all p > 0.05).

There was no significant seasonal trend in FOG levels, as shown by the
non-significance of a quadratic model of the dependence of recorded FOG level
on the day of year or that record. Similarly, there was no significant
relationship with temperature (all p > 0.05).

```{r fogcont, echo=FALSE, fig.pos='h', fig.height=3.5, fig.cap=sprintf('The highest number of blockages (shown by dark areas on the contour plot) occurs when levels of FOG are high, and air temperatures are low.'), message=FALSE, eval=.do.grid}
if (.do.grid) {
    plot(.plot.grid)
}
```


### Model validation

The response of weekly grease blockage frequency to air temperature was fit
using observations prior to `r block.airtemp.pred$cutoff` 
(`r nrow(block.airtemp.pred$dat)` weeks).  The resulting model was used to predict
grease blockage frequency in the weeks after `r block.airtemp.pred$cutoff` 
(`r nrow(block.airtemp.pred$pred)` weeks).  The results, shown in 
Figure \@ref(fig:validate), indicate that model predictions generally capture the
observed pattern of grease blockage frequency, though with a tendency to
underestimate blockages during weeks with the highest frequencies.

```{r validate, echo=FALSE, fig.pos='h', fig.height=3.5, fig.cap=sprintf('Model validation comparing predicted and observed weekly grease blockage frequency. To generate predictions, a model was fit using weeks prior to %s (%s weeks total).  Observed air temperatures in subsequent weeks (%s weeks total) were used to predict grease blockage frequency (X-axis). Also plotted are the observed grease blockage frequencies in those weeks (Y-axis). The dotted line is the identity line y = x, and the solid line shows the linear relationship between observed and predicted blockages (y = %sx %s).', block.airtemp.pred$cutoff, nrow(block.airtemp.pred$dat), nrow(block.airtemp.pred$pred), round(coef(block.airtemp.pred$pred.lm), 2)[2], round(coef(block.airtemp.pred$pred.lm), 2)[1] ), message=FALSE}
plot(block.airtemp.pred$plot)
```

### Holiday FOG?

Using residuals from the model of blockage frequency by air temperature and
cause, we tested the hypothesis that the holiday period causes extra blockages
via the addition of extra fats and grease to the drainage system.  
Figure \@ref(fig:tday) compares model residuals from the weeks containing and
immediately after Thanksgiving with other weeks, isolating any difference in
blockage frequency during this period once temperature is taken into account.
No difference is observed between the holiday period and other times.

```{r tday, echo=FALSE, fig.pos='h', fig.height=3.5, fig.cap=sprintf('Once air temperature is accounted for, the weeks containing and following Thanksgiving show no evidence of elevated frequencies of grease-caused blockages. Points are horizontally jittered to avoid overlapping.'), message=FALSE}
plot(block.airtemp.list$tday.plot)
```

## Discussion

We have demonstrated a significant, conserved pattern of FOG-caused blockages
in response decreased environmental temperatures in the city of Albuquerque, New
Mexico. We do not explicitly test the mechanisms driving these patterns; nonetheless,
our findings are broadly consistent with an extensive body of literature on 
FOG blockage formation. The city of Albuquerque has a number of unique geographic and
climactic features, including large yearly temperature swings, highly variable
precipitation, a significant within-system elevation gradient, and extensive
cold air drainage.  Whether our results will hold in other locales remains an
intriguing question. The ready availability of automated air temperature
records suggests that the analyses presented here could easily be replicated in
other cities where sanitary sewer blockage data is routinely collected.  A
comparison between cities from a range of climates and topographies could
provide valuable evidence on empirical patterns of FOG blockages.

We find that mean air temperature has only modest utility in predicting sewer
blockages over weekly timescales. Nonetheless, an increased understand of the
drivers of SSOs in general, and FOG-related blockages in particular, could aid
in system design and maintainence.  Consideration could be given in sewer
design to select locations subject to higher winter temperatures, either
through greater depth or under asphalt pavement.  Similarly, cleaning
operations could be prioritized to areas expected to experience coldest
temperatures.

Our results suggest that sewer temperature provides a somewhat more accurate
predictor of blocking frequency than air temperature. In systems where these
data are regularly collected and made available to managers, such data could
potentially be used anticipate problems in sanitation infrastructure and plan
system maintenance.

We note that sewage temperature is driven primarily by soil temperatures
(author’s communication with Dr. Jes Vollertsen, 2014), and a body of
literature on the prediction of soil temperature from air temperature in
non-urban areas exists (e.g., @Hasfurther1972). The relationship between air
and soil temperature is likely to be mediated by groundwater levels.
Differences in the response of sewer temperature to air temperature between
interceptors and manholes likely reflects elevation and land use, possibly via
their effects on groundwater temperature.  Models including precipitation
patterns and/or local physical characteristics (e.g. water table height, land
use, sewer configuration, soil type, geology) could test this hypothesis. The
existence of within-system variation in this study suggest that between-system
variation may also be significant.

Seasonal onset of lower environmental temperatures commonly coincide with 
the Thanksgiving and Christmas season in the Northern Hemisphere. As such, 
lower temperatures are a possible explanation of the popular association of 
FOG related blockages with the holiday-related activities.  
At the system-level, however, we find no excess of blockages during holiday 
weeks above those expected given prevailing temperatures. 
Further study of SSO rates versus holiday seasons in
Southern Hemisphere cities could test this hypothesis.

The municipal system described here is a separate (SSS) rather than a combined 
sanitary sewer system (CSS). As noted, SSS are common in the U.S., whereas CSS
are more common in regions with higher average infrastructure age (e.g., Europe)
[@burian2000urban].
Furthermore, the U.S. has a relatively high prevalence of garbage disposal units, i.e., 
mechanical grinders that add food waste directly into the sewer system [@iacovidou2012household]. 
The lack of storm flows and the addition of ground food waste
may contribute to the observed findings.  As such, an important test of our
findings is to asses the impact of temperature on blockage frequency in regions
such as Europe where combined sewers are more common, and garbage disposal
units rarer. 

[](Comment = empty link)
[](There may also be differences between old and new sewer systems.)

Our results also suggest that areas experiencing increasing average
temperatures due to climate change may experience an overall reduction in
blockage frequencies over time.  Indeed, large-scale climatic trends could
serve as natural experiments to study the impact of temperature on long-term
(e.g. yearly) blockage frequencies.

With continuing population rise and urbanisation, efficient operation of urban
waste-water infrastructure is an increasingly important issue for global public
health.  @Sato2013 recently highlighted the importance of more research into
efficacy of waste-water treatment techniques, particularly in the developing
world. The interaction between water conservation measures and sanitary sewer 
performance, for example, can place conflicting demands on municipalities 
[@marleni2012impact]. 
The sewer blockage and temperature data presented here were collected
during routine system monitoring in the course of standard operations.  
We hope this work demonstrates the potential of historic datasets to
address modern and emerging urban infrastructure challenges.

## Tables

```{r tempandblock, echo=FALSE, results='asis'}
.cap <- with(block.sewtemp.list, sprintf(
    'Weekly mean sewage temperature and blockage cause predict weekly blockage frequency ( NB-GLM, n=%d weeks, D=%s)', 
    nweeks, dev
))
ret <- my.xtable(block.sewtemp.list$mod,
    specs= ~variable, caption=.cap
)
print(ret)
```

```{r greasetab, echo=FALSE, results='asis'}
.cap <- with(block.airtemp.list, sprintf(
    'Weekly mean air temperature and blockage cause predicts weekly blockage frequency ( NB-GLM, n=%d weeks, D=%s)', 
    nweeks, dev
)) 
my.xtable(block.airtemp.list$mod, specs= ~variable, caption=.cap)
```
